---
title: "STAT 545B Assignment B1 Deliverable"
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

# Exercise 1 & 2: Make and Document a Function    

One example of a versatile series of functions is to `group_by()` one or more variables, then calculate the proportion and count of a specific observation within these groups. 

## 1.0 Examples of Grouping & Counting Conditional Observations

```{r}
#Loading the necessary packages
library(tidyverse) #includes dplyr package for data wrangling, used in designed function
library(palmerpenguins) #contains sample data set
library(gapminder) #contains sample data set
library(testthat) #includes multiple formal tests for function behaviours
```


### 1.0.1 `penguins` Data Set  

```{r}
#Previewing `penguins` data set 
head(penguins)
```

```{r}
#Summarizing the variables of `penguins` data set 
summary(penguins)
```

The following code groups the above data by `species` and `island`, then counts the number and proportion of observations where `bill_length_mm > 40`.

```{r}
penguins %>% group_by(species, island) %>%
  summarize(n = n(),
            count = length(which(bill_length_mm > 40)),
            prop = (length(which(bill_length_mm > 40))/n())*100)
```

### 1.0.2 `gapminder` Data Set

```{r}
#Previewing `gapminder` data set 
head(gapminder)
```

```{r}
#Summarizing the variables of `gapminder` data set 
summary(gapminder)
```

The following code groups the above data by `year`, then counts the number and proportion of observations where `lifeExp > 60`.

```{r}
gapminder %>% group_by(year) %>%
  summarize(n = n(),
            count = length(which(lifeExp > 60)),
            prop = (length(which(lifeExp > 60))/n())*100)
```


## 2.0 Designing a Generic Function  

The following code creates a function called `group_count_and_prop` that uses three inputs: `.data`, `...` and `cond` to call (1) a data frame, (2) variables to group by and (3) the condition to evaluate, respectively. The output is a new data frame summarizing the number and proportion of observations in the input data frame that match the input condition within each given group. 

```{r}
#' @title Count and calculate proportions in group(s) 
#' 
#' @description Summarizes the number and proportion of observations that meet one or more conditions within given group(s).
#' 
#' @param .data A data frame, data frame extension (e.g. a tibble), or a lazy data frame (e.g. from dbplyr or dtplyr).
#' @param ... Variables to group by.
#' @param cond Short for "condition". Expressions that return a logical value. Expressions are defined in terms of variables in .data and multiple expressions, if included, must be combined with the & operator. Only rows for which all conditions return TRUE are evaluated.  
#' 
#' @return A new data frame summarizing the groups, total number of observations in each group, count of observations that return TRUE for the condition(s) and proportion (%) of observations that return TRUE for the condition(s)
#' @export
#' 
#' @examples
#' group_count_and_prop(starwars, gender, cond = height > 150 & mass > 50)
#' group_count_and_prop(mtcars, cyl, cond = mpg > 20)

group_count_and_prop <- function(.data, ..., cond) {
  .data %>% group_by(...) %>%
    summarize(n = n(),
              count = length(which({{ cond }})),
              prop = (length(which({{ cond }}))/n())*100)
}
```
  
  
# Exercise 3: Include Examples

## 3.0 Examples

### 3.0.1 `penguins` Data Set  
Recall the `penguins` data set introduced in **Section 1.0**.

```{r}
#Summarizing the variables of `penguins` data set
summary(penguins)
```

Here is an example of employing the `group_count_and_prop` function to group the `penguins` data set by `species` and `island`, then counting the number and proportion of observations where `bill_length_mm > 40`. The resulting tibble matches the output generated in **Section 1.0**.

```{r}
group_count_and_prop(penguins,
                     species, island,
                     cond = bill_length_mm > 40)
```

Here is an an example of employing the `group_count_and_prop` function to group the `penguins` data set by `species`, then counting the number and proportion of observations where `body_mass_g > 4000` and `sex == "female"`. 

```{r}
group_count_and_prop(penguins,
                     species,
                     cond = body_mass_g > 4000 & sex == "female")
```

  
### 3.0.2 `gapminder` Data Set  
Recall the `gapminder` data set introduced in **Section 1.0**.

```{r}
#Summarizing the variables of `gapminder` data set
summary(gapminder)
```

Here is an example of employing the `group_count_and_prop` function to group the `gapminder` data set by `year`, then counting the number and proportion of observations where `lifeExp > 60`. The resulting tibble matches the output generated in **Section 1.0**.

```{r}
group_count_and_prop(gapminder, 
                     year,
                     cond = lifeExp > 60)
```

  
# Exercise 4: Test the Function  
The following section formally tests the `group_count_and_prop` function given various types of input to test the robustness of its intended use and its response to inappropriate inputs.

## 4.0 Multiple inputs in single arugment 
These scenarios test the flexibility of the two variable inputs in `group_count_and_prop` for the embedded `group_by` and `length(which())` functions. 

### 4.0.1 Vector with no NAs, **two** `group_by` inputs  
The following tibble is the output of the `gapminder` data set grouped by `continent` and `year`, followed by the count and proportion of observations where `lifeExp > 60`.

```{r}
#Expected output when ... = continent, year, cond = lifeExp > 60 in `gapminder`
test401 <- gapminder %>% group_by(continent, year) %>%
  summarize(n = n(),
            count = length(which(lifeExp > 60)),
            prop = (length(which(lifeExp > 60))/n())*100)

print(test401)
```

This test evaluates if the outputs of `group_count_and_prop` versus `test401` are identical.

```{r}
#Test function against expected outcome test401
test_that("returns vector grouped by two variables, count and prop by each group", {
  expect_identical(object = group_count_and_prop(gapminder, 
                                                 continent, year,
                                                 cond = lifeExp > 60),
                   expected = test401)
})
```

### 4.0.2 Vector with no NAs, **two** `cond` inputs  
The following tibble is the output of the `penguins` data set grouped by `species`, followed by the count and proportion of observations where `flipper_length_mm > 200` and `sex == "male"`.

```{r}
#Expected output when ... = species, cond = flipper_length_mm > 200 & sex == "male" in `penguins`
test402 <- penguins %>% group_by(species) %>%
  summarize(n = n(),
            count = length(which(flipper_length_mm > 200 & sex == "male")),
            prop = (length(which(flipper_length_mm > 200 & sex == "male"))/n())*100)

print(test402)
```

This test evaluates if the outputs of `group_count_and_prop` versus `test402` are identical.

```{r}
#Test function against expected outcome test402
test_that("returns vector with count and prop of observations matching two conditions", {
  expect_identical(object = group_count_and_prop(penguins, 
                                                 species,
                                                 cond = flipper_length_mm > 200 & sex == "male"),
                   expected = test402)
})
```

### 4.1 Invalid `group_by` argument  
The following code returns an error due to the `group_by` argument not being found in the data frame, `penguins`. 

```{r, error = TRUE}
#Expected error given a group_by() input not found in `penguins`
penguins %>% group_by(notfound) %>%
  summarize(n = n(),
            count = length(which(body_mass_g > 4000)),
            prop = (length(which(body_mass_g > 4000))/n())*100)
```

This test evaluates if inputting the same invalid `group_by` argument in `group_count_and_prop` returns an error.

```{r}
#Test function with invalid group_by() argument
test_that("error given invalid group_by argument", {
  expect_error(object = group_count_and_prop(penguins,
                                             notfound,
                                             cond = body_mass_g > 4000))
})
```

### 4.2 Invalid `cond` argument  
The following code returns an error due to a string input to `which()` when a logical expression is required. 

```{r, error = TRUE}
#Expected error given a string input in which() argument
gapminder %>% group_by(continent) %>%
  summarize(n = n(),
            count = length(which("notlogical")),
            prop = (length(which("notlogical"))/n())*100)
```

This test evaluates if inputting the same invalid `cond` argument in `group_count_and_prop` returns an error.

```{r}
#Test function with invalid `cond` argument
test_that("error given invalid cond argument", {
  expect_error(object = group_count_and_prop(gapminder,
                                             continent,
                                             cond = "notlogical"))
})
```

