---
title: "STAT 545B Assignment B1 Deliverable"
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

# Exercise 1 & 2: Make and Document a Function    

One example of a versatile series of functions is to `group_by()` one or more variables, then calculate the proportion and count of a specific observation within these groups. 

## 1.0 Examples of Grouping & Counting Conditional Observations

```{r}
#Loading the necessary packages
library(tidyverse)
library(palmerpenguins)
library(gapminder)
library(devtools)
library(roxygen2)
```


### 1.1 `penguins` Data Set  

```{r}
#Previewing `penguins` data set 
head(penguins)
```

```{r}
#Summarizing the variables of `penguins` data set 
summary(penguins)
```

The following code groups the above data by `species` and `island`, then counts the number and proportion of observations where `bill_length_mm > 40`.

```{r}
penguins %>% group_by(species, island) %>%
  summarize(n = n(),
            count = length(which(bill_length_mm > 40)),
            prop = (length(which(bill_length_mm > 40))/n())*100)
```

### 1.2 `gapminder` Data Set

```{r}
#Previewing `gapminder` data set 
head(gapminder)
```

```{r}
#Summarizing the variables of `gapminder` data set 
summary(gapminder)
```

The following code groups the above data by `year`, then counts the number and proportion of observations where `lifeExp > 60`.

```{r}
gapminder %>% group_by(year) %>%
  summarize(n = n(),
            count = length(which(gdpPercap > 9000)),
            prop = (length(which(gdpPercap > 9000))/n())*100)
```


## 2.0 Designing a Generic Function  

The following code creates a function called `group_count_and_prop` that uses three inputs: `.data`, `...` and `.cond` to call a data frame, variables to group by and the condition to count, respectively. The output is a new data frame summarizing the number and proportion of observations in the input data frame that match the input condition within each given group. 

```{r}
#' @title Count and calculate proportions in group(s) 
#' @description Summarizes the number and proportion of observations that meet one or more conditions within given group(s).
#' @param .data A data frame, data frame extension (e.g. a tibble), or a lazy data frame (e.g. from dbplyr or dtplyr).
#' @param ... Variables to group by.
#' @param cond Short for "condition". Expressions that return a logical value. Expressions are defined in terms of variables in .data and multiple expressions, if included, must be combined with the & operator. Only rows for which all conditions return TRUE are evaluated.  
#' @return A new data frame summarizing the groups, total number of observations in each group, count of observations that return TRUE for the condition(s) and proportion (%) of observations that return TRUE for the condition(s)
#' @examples
#' group_count_and_prop(starwars, gender, cond = height > 150 & mass > 50)
#' group_count_and_prop(mtcars, cyl, cond = mpg > 20)
#' @export

group_count_and_prop <- function(.data, ..., cond) {
  .data %>% group_by(...) %>%
    summarize(n = n(),
              count = length(which({{ cond }})),
              prop = (length(which({{ cond }}))/n())*100)
}
```
  
  
# Exercise 3: Include Examples

## 3.0 Examples

### 3.1 `penguins` Data Set  
Recall the `penguins` data set introduced in Section 1.0.

```{r}
#Summarizing the variables of `penguins` data set
summary(penguins)
```

Here is an example of employing the `group_count_and_prop` function to group the `penguins` data set by `species` and `island`, then counting the number and proportion of observations where `bill_length_mm > 40`. The resulting tibble matches the output generated in Section 1.0.

```{r}
group_count_and_prop(penguins,
                     species, island,
                     cond = bill_length_mm > 40)
```

Here is an an example of employing the `group_count_and_prop` function to group the `penguins` data set by `species`, then counting the number and proportion of observations where `body_mass_g > 4000` and `sex == "female"`. 

```{r}
group_count_and_prop(penguins,
                     species,
                     cond = body_mass_g > 4000 & sex == "female")
```

### 3.2 `gapminder` Data Set  
Recall the `gapminder` data set introduced in Section 1.0.

```{r}
#Summarizing the variables of `gapminder` data set
summary(gapminder)
```

Here is an example of employing the `group_count_and_prop` function to group the `gapminder` data set by `year`, then counting the number and proportion of observations where `lifeExp > 60`. The resulting tibble matches the output generated in Section 1.0.

```{r}
group_count_and_prop(gapminder, 
                     year,
                     cond = lifeExp > 60)
```

# Exercise 4: Test the Function

